---
title: "What disease has caused the highest deaths in the world before the COVID-19 pandemic?"
description: |
  An analysis on the mortality rate for each caused of deaths category.
author: "Krisanat Anukarnsakulchularp"
date: 2022-09-06
categories: [mortality, R]
format:
  html:
    toc: true
    toc_float: true
    fig-align: center
    number-sections: false
    theme: solar
---

```{r setup}
#| include: false

knitr::opts_chunk$set(echo = FALSE,
                      eval = TRUE,
                      message = FALSE,
                      warning = FALSE,
                      error = FALSE)
```

```{r}
# Loading library
library(tidyverse)
library(tmap)
library(sf)
library(RColorBrewer)
library(knitr)
library(kableExtra)
```

```{r data}
# Mortality data
mortality <- read_csv(("data/share-of-deaths-by-cause.csv")) %>%
  select(Entity,
         Code,
         Year,
         `Deaths - Cardiovascular diseases - Sex: Both - Age: All Ages (Percent)`,
         `Deaths - Neoplasms - Sex: Both - Age: All Ages (Percent)`,
         `Deaths - Maternal disorders - Sex: Both - Age: All Ages (Percent)`,
         `Deaths - Chronic respiratory diseases - Sex: Both - Age: All Ages (Percent)`,
         `Deaths - Digestive diseases - Sex: Both - Age: All Ages (Percent)`,
         `Deaths - Diabetes mellitus - Sex: Both - Age: All Ages (Percent)`,
         `Deaths - Lower respiratory infections - Sex: Both - Age: All Ages (Percent)`,
         `Deaths - Neonatal disorders - Sex: Both - Age: All Ages (Percent)`,
         `Deaths - Diarrheal diseases - Sex: Both - Age: All Ages (Percent)`,
         `Deaths - Cirrhosis and other chronic liver diseases - Sex: Both - Age: All Ages (Percent)`,
         `Deaths - Tuberculosis - Sex: Both - Age: All Ages (Percent)`,
         `Deaths - Chronic kidney disease - Sex: Both - Age: All Ages (Percent)`,
         `Deaths - Alzheimer's disease and other dementias - Sex: Both - Age: All Ages (Percent)`,
         `Deaths - Parkinson's disease - Sex: Both - Age: All Ages (Percent)`,
         `Deaths - HIV/AIDS - Sex: Both - Age: All Ages (Percent)`,
         `Deaths - Acute hepatitis - Sex: Both - Age: All Ages (Percent)`,
         `Deaths - Malaria - Sex: Both - Age: All Ages (Percent)`,
         `Deaths - Nutritional deficiencies - Sex: Both - Age: All Ages (Percent)`,
         `Deaths - Meningitis - Sex: Both - Age: All Ages (Percent)`,
         `Deaths - Protein-energy malnutrition - Sex: Both - Age: All Ages (Percent)`,
         `Deaths - Enteric infections - Sex: Both - Age: All Ages (Percent)`) %>%
  mutate(Code = toupper(Code))

# Map data
world_map <- read_sf(("data/world-administrative-boundaries/world-administrative-boundaries.shp")) %>%
  st_zm(drop = TRUE) %>%
  mutate(iso3 = toupper(as.character(iso3)))

# Extract countries code
countries <- pull(world_map, iso3)

# Joining data set (containing only countries)
mortality_sf <- mortality %>%
  filter(toupper(Code) %in% countries) %>%
  group_by(Code) %>%
  right_join(world_map, by = c("Code" = "iso3")) %>%
  ungroup()

# Clean format
clean_mortality <- mortality_sf %>%
  select(Entity,
         Code,
         Year,
         `Deaths - Cardiovascular diseases - Sex: Both - Age: All Ages (Percent)`,
         `Deaths - Neoplasms - Sex: Both - Age: All Ages (Percent)`,
         `Deaths - Maternal disorders - Sex: Both - Age: All Ages (Percent)`,
         `Deaths - Chronic respiratory diseases - Sex: Both - Age: All Ages (Percent)`,
         `Deaths - Digestive diseases - Sex: Both - Age: All Ages (Percent)`,
         `Deaths - Diabetes mellitus - Sex: Both - Age: All Ages (Percent)`,
         `Deaths - Lower respiratory infections - Sex: Both - Age: All Ages (Percent)`,
         `Deaths - Neonatal disorders - Sex: Both - Age: All Ages (Percent)`,
         `Deaths - Diarrheal diseases - Sex: Both - Age: All Ages (Percent)`,
         `Deaths - Cirrhosis and other chronic liver diseases - Sex: Both - Age: All Ages (Percent)`,
         `Deaths - Tuberculosis - Sex: Both - Age: All Ages (Percent)`,
         `Deaths - Chronic kidney disease - Sex: Both - Age: All Ages (Percent)`,
         `Deaths - Alzheimer's disease and other dementias - Sex: Both - Age: All Ages (Percent)`,
         `Deaths - Parkinson's disease - Sex: Both - Age: All Ages (Percent)`,
         `Deaths - HIV/AIDS - Sex: Both - Age: All Ages (Percent)`,
         `Deaths - Acute hepatitis - Sex: Both - Age: All Ages (Percent)`,
         `Deaths - Malaria - Sex: Both - Age: All Ages (Percent)`,
         `Deaths - Nutritional deficiencies - Sex: Both - Age: All Ages (Percent)`,
         `Deaths - Meningitis - Sex: Both - Age: All Ages (Percent)`,
         `Deaths - Protein-energy malnutrition - Sex: Both - Age: All Ages (Percent)`,
         `Deaths - Enteric infections - Sex: Both - Age: All Ages (Percent)`,
         continent,
         geometry) %>%
  na.omit() %>%
  pivot_longer(-c(Entity, Code, Year, continent, geometry),
               names_to = "Disease",
               values_to = "Mortality_rate_(%)") %>%
  mutate(Disease = str_replace_all(Disease, c("Deaths - " = "", " - Sex: Both - Age: All Ages \\(Percent\\)" = "")),
         `Mortality_rate_(%)` = round(`Mortality_rate_(%)`, 4)) %>%
  relocate(c(continent, Year, Disease, `Mortality_rate_(%)`, geometry), .after = c(Code, continent, Year, Disease, `Mortality_rate_(%)`))
```

```{r}
# Clean
clean_mortality_highest <- clean_mortality %>%
  group_by(Entity, Year) %>%
  mutate(Highest = max(`Mortality_rate_(%)`),
         Cause = ifelse(Highest == `Mortality_rate_(%)`, "Yes", "No")) %>%
  filter(Cause == "Yes") %>%
  select(-c(Cause, Highest))

clean_mortality_highest <- st_as_sf(clean_mortality_highest) %>%
  filter(Year == 2019)

disease_name <- unique(clean_mortality$Disease)
country_name <- unique(clean_mortality$Entity)
continent_name <- unique(clean_mortality$continent)

disease_ordered <- clean_mortality_highest %>%
  group_by(Disease) %>%
  summarise(n = n()) %>%
  arrange(-n) %>%
  pull(Disease)

clean_mortality_highest$Disease <- ordered(clean_mortality_highest$Disease, levels = disease_ordered)

my_palette <- rev(brewer.pal(name = "OrRd", n = 8))
my_palette <- my_palette[-2]
```


In 2019, there are about 55.4 million deaths worldwide, which accounted for approximately 55% of the population (“The Top 10 Causes of Death”, 2020).

Since the COVID-19 pandemic, multiple countries have shifted most of their focus and resources to fight against the virus. This has shadowed many other existing diseases. In this blog post, I will use mortality data from [Our World in Data](https://ourworldindata.org/causes-of-death) to look at the main diseases that impact the population.


## Data description

The mortality data set contains information about the causes of death. The variables that will be used in this blog are the following:

-   Entity: This contains country name, continent, and different income level

-   Code: The country code stored in the ISO3 format

-   Year: Time period from 1990 - 2019

-   Causes of deaths: The mortality rate for each cause of deaths type (21 diseases types, 11 other types)

For the cause of death data, I will only use the disease information, and other causes of death that are not related to diseases will not be used. Also, the **Entity** variable will only contain the country name. The continent and different income levels will be removed. 

To ensure that the data can be visualized using the map, the map data set obtained from [opendatasoft](https://public.opendatasoft.com/explore/dataset/world-administrative-boundaries/export/) has been included. The *read_sf* function will be used to read the shapefile data. I will only be using the following variable:

-   iso3: The ISO 3 territory code

-   continent: The continent of the territory

-   geometry: sf object contains the boundary for each territory

The following alteration has been implemented:

-   Using the *select* function, only the causes of deaths that are health-related are selected

-   Using the *filter* function, filtered the **Entity** to only contain the countries (removing other entities, e.g., world bank lower income)

-   Using the *right_join* function to join the mortality data set and world map together

-   Using the *pivot_longer* function to put all the causes of death variables into one column called **Disease** and the mortality rate into the column called **Mortality_rate_(%)**


## Highest cause of death in each country

@fig-worldmap shows that the disease that causes most death is Cardiovascular disease, as the deepest red color represents the highest number of countries that are affected by the disease. The second disease that causes the most death is Neoplasms (also known as cancer).

```{r}
#| label: fig-worldmap
#| fig-cap: "Highest cause of death in each country"

clean_mortality_highest %>%
  ggplot(aes(fill = Disease)) +
  geom_sf() +
  scale_fill_manual(values = my_palette) +
  theme_bw() +
  labs(title = "Highest Caused of death for each country in 2019") +
  theme(legend.position = "bottom", 
        legend.title = element_text(size = 12, face = "bold"), 
        plot.title = element_text(size = 12, face = "bold"))
```

From @fig-worldmap, I have found some interesting points. Cardiovascular diseases highly affected most countries in the world. However, Australia and countries in Africa are affected by other diseases, such as neoplasms and HIV/AIDS. It is because the income level in Africa is low, which means less access to a good education leads to less information about communicable diseases like HIV/AIDS. In Australia, an increase in mortality rate from neoplasms are due to the growth in size and aging of the Australian population ("Cancer burden and control in Australia: lessons learnt and challenges remaining", 2018).


## Top 10 higest mortality rate in the world for 2019

There are two main types of disease, Non-communicable and Communicable diseases. Non-communicable diseases (NCDs) it is a disease that can not be passed on from person to person, while communicable diseases can spread from person to person. Now let’s see the types of diseases in 2019 that are the most deadly in the world.

<div>

<a name = 'top10'></a>

```{r}
clean_mortality %>%
  filter(Year == 2019) %>%
  group_by(Disease) %>%
  summarise(Mortality_rate = mean(`Mortality_rate_(%)`), .groups = "drop") %>%
  arrange(-Mortality_rate) %>%
  select(-Mortality_rate) %>%
  mutate(Rank = paste0(1:21,".")) %>%
  relocate(Disease, .after = Rank) %>%
  head(10) %>%
  kable(caption = "Top 10 highest mortality rate in the world for 2019") %>%
  kable_styling(bootstrap_options = c("striped", "hover"), full_width = FALSE)
```

</div>

From the [top-10 table](#top10), there are seven diseases that are NCD. Cardiovascular diseases, Neoplasms, Chronic respiratory diseases, Diabetes mellitus, Chronic kidney disease, Alzheimer's disease, Cirrhosis, and other chronic liver diseases. While the two diseases, Lower respiratory infections and Neonatal disorders classified as Communicable diseases. Digestive diseases can be both NCD and CD.

As there are up to seven out of ten NCDs in this global list, it shows that the impact of NCDs is larger than how it is currently portrayed in society nowadays. This implies that people should be more aware of the consequences that come from these diseases. Also, from medical professionals' perspective, investment in more effective treatment should be made one of the main priorities. 


## Impact of NCD over time

To further analyze the impact of NCD, I will plot the time series for each disease and see how it impacts over time.

```{r}
#| label: fig-NCD
#| fig-cap: "Impact of NCD over time"
#| fig-height: 8

clean_mortality %>%
  filter(Disease %in% c("Cardiovascular diseases",
                        "Neoplasms",
                        "Chronic respiratory diseases",
                        "Diabetes mellitus",
                        "Chronic kidney disease",
                        "Alzheimer's disease and other dementias",
                        "Cirrhosis and other chronic liver diseases")) %>%
  group_by(Disease, Year) %>%
  summarise(Mortality_rate = mean(`Mortality_rate_(%)`), .groups = "drop") %>%
  ggplot(aes(x = Year, y = Mortality_rate, color = Disease)) +
  geom_line() +
  geom_point() +
  labs(x = "Year", y = "Mortality Rate (percentage)") +
  theme_bw() +
  theme(legend.position = "bottom", 
        legend.title = element_text(size = 12, face = "bold"), 
        plot.title = element_text(size = 12, face = "bold"),
        axis.text.x = element_text(angle = 50, vjust = 1, hjust = 1),
        strip.text = element_text(size = 7)) +
  facet_wrap(~Disease, scales = "free")
```

From @fig-NCD the trend of the mortality rate seems to increase over time, which is very concerning as this shows that there is not enough effort put into developing the treatment needed for these diseases. 

As media continues to shift focus to the new communicable diseases like COVID-19 and Monkey Pox, it is still vital for people around the world to recognize how deadly some common diseases can be. For the global population, having access to accurate and educational information about these diseases is important as it will allow people to look out for the symptoms and protect themselves effectively. On the other hand, from the medical point of view, as earlier mentioned before, there should be more resources such as more accessibility to treatment and medicines for these diseases. The implementation of these recommendations can potentially be beneficial for not only medical professionals but for many populations around the world as well.


## References

Cheng, E. S., Weber, M., Feletto, E., Smith, M. A., & Yu, X. Q. (2018). Cancer burden and control in Australia: lessons learnt and challenges remaining. Annals of Cancer Epidemiology,
  2(0). https://ace.amegroups.com/article/view/4614/5363

Neuwirth E (2022). _RColorBrewer: ColorBrewer Palettes_. R package version 1.1-3, <https://CRAN.R-project.org/package=RColorBrewer>.

Pebesma, E., 2018. Simple Features for R: Standardized Support for Spatial Vector Data. The R Journal 10 (1), 439-446, https://doi.org/10.32614/RJ-2018-009

Ritchie, H., & Roser, M. (2018, February). Causes of Death. Retrieved from Our World in Data website: https://ourworldindata.org/causes-of-death

Tennekes M (2018). “tmap: Thematic Maps in R.” _Journal of Statistical Software_, *84*(6), 1-39. doi:10.18637/jss.v084.i06 <https://doi.org/10.18637/jss.v084.i06>.

Wickham H, Averick M, Bryan J, Chang W, McGowan LD, Françoi R, Grolemun G, Haye A, Henr L, Heste J, Kuh M, Pederse TL, Mille E, Bach SM, Müll K, Oo ,J, Robins ,D, Seid ,DP, Spi ,V,
  Takahas ,K, Vaugh ,D, Wil ,C, W ,K, Yutani ,H (2019). “Welcome to the tidyverse.” _Journal of Open Source Software_, *4*(43), 1686. doi:10.21105/joss.01686
  <https://doi.org/10.21105/joss.01686>.

World Administrative Boundaries - Countries and Territories. (n.d.). Retrieved September 6, 2022, from public.opendatasoft.com website:
  https://public.opendatasoft.com/explore/dataset/world-administrative-boundaries/export/

World Health Organization. (2020, December 9). The Top 10 Causes of Death. Retrieved from World Health Organization website:
  https://www.who.int/news-room/fact-sheets/detail/the-top-10-causes-of-death

Yihui Xie (2022). knitr: A General-Purpose Package for Dynamic Report Generation in R. R package version 1.39.

Yihui Xie (2015) Dynamic Documents with R and knitr. 2nd edition. Chapman and Hall/CRC. ISBN 978-1498716963

Yihui Xie (2014) knitr: A Comprehensive Tool for Reproducible Research in R. In Victoria Stodden, Friedrich Leisch and Roger D. Peng, editors, Implementing Reproducible
  Computational Research. Chapman and Hall/CRC. ISBN 978-1466561595

Zhu H (2021). _kableExtra: Construct Complex Table with 'kable' and Pipe Syntax_. R package version 1.3.4, <https://CRAN.R-project.org/package=kableExtra>.


